Variational theory of non-relativistic quantum electrodynamics

### 2019-02-06 paper4pg.tex
small corrections and getting rid of few more words

\begin{abstract}
The ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and new insights into quantum electrodynamics (QED). Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems. In this Letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter. Essential to our ansatz is the notion of an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling. This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space. As a first step towards a new  \textit{ab initio} approach to ground and excited state energies in QED, we apply our ansatz to describe a multi-level emitter coupled to many optical modes, a system with no analytical solution. We find a compact semi-analytical formula which describes ground and excited state energies very well in all regimes of coupling parameters allowed by sum rules. Our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity. Our method should thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems. 
\end{abstract}

@@

In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles. Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation. A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation. Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom. A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT. 
 %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

@@

The QED Hamiltonian is given by $H = H_{\mathrm{mat}}+H_{\mathrm{em}}+H_{\mathrm{int}}$ where $H_{\mathrm{mat}}$ describes the matter in the absence of the quantized electromagnetic field, $H_{\mathrm{em}}$ describes the photons in the absence of the matter, and $H_{\mathrm{int}}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:
\begin{align}
H_{\mathrm{el}} &= \int d^3x ~\psi^{\dagger}(\mathbf{x})\left(-\frac{\hbar^2\nabla^2}{2m} + v_{\mathrm{ext}}(\mathbf{x}) \right)\psi(\mathbf{x}) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(\mathbf{x})\psi^{\dagger}(\mathbf{x}')V(\mathbf{x}-\mathbf{x}')\psi(\mathbf{x}')\psi(\mathbf{x}),
\end{align}
where $v_{\mathrm{ext}}$ is the one-body external potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
\begin{figure}[t]
\includegraphics[width=8.5cm]{conceptfigure.pdf}
\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons. (Right) Quasiparticle description of the coupled system as a factorizable state of an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
\label{fig:ansatz}
\end{figure}
Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$ renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{\mathrm{em}} = \frac{\epsilon_0}{2}\int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity.

@@

We capture the effect of correlations perturbatively. For the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{\infty}\sum_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
\label{eq:e-shift}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, and$N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs. (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq. (\ref{eq:e-shift}). By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found. \textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq. ~\ref{eq:e-shift}. These equations take into account the spatially varying wavefunctions to the spatially varying mode functions, just like Eqs. \ref{eq:mhf-electron} and \ref{eq:mhf-photon}, and therefore do not assume the dipole approximation.}
%\begin{figure}[t]
%\includegraphics[width=8.4cm]{figure3new.pdf}
%\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
%\label{fig:ansatz}
%\end{figure}

@@

Importantly, the reason perturbation theory fails for first excited state is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels. In contrast, no such near-resonance is found in the variational framework because the plasma term  in Eq. (\ref{eq:mhf-photon}) blue-shifts all of the photon frequencies. For the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV (from 0.62 eV),  far off-resonance from the bare emitter transition. The plasma term also strongly reduces the coupling between light and matter by a mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field (see \ref{fig:results}b). This is a  light-matter decoupling effect, which was proposed in Ref. ~\cite{liberato2014}. \textcolor{blue}{In Ref. ~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, deLiberato obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}

\textcolor{blue}{This light-matter decoupling is also reflected in Fig. ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials). As shown in Fig. 3 be seen, numerical and variational methods capture a saturation and then decrease of this expectation value. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}

>>>>

\begin{abstract}
The ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and new insights into quantum electrodynamics (QED). Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems. In this Letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter. Essential to our ansatz is the notion of an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling. This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space. As a first step towards a new  \textit{ab initio} approach to ground and excited state energies in QED, we apply our ansatz to describe a multi-level emitter coupled to many optical modes, a system with no analytical solution. We find a compact semi-analytical formula which describes ground and excited state energies very well in all regimes of coupling parameters allowed by sum rules. Our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity. Our method thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems. 
\end{abstract}

@@

In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles. Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation. A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation. Beyond these advantages, a variational framework also allows conceptual insights, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom. A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT. 
 %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

@@

In general, the QED Hamiltonian is given by $H = H_{\mathrm{mat}}+H_{\mathrm{em}}+H_{\mathrm{int}}$ where $H_{\mathrm{mat}}$ describes the matter in the absence of the quantized electromagnetic field, $H_{\mathrm{em}}$ describes the photons in the absence of the matter, and $H_{\mathrm{int}}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:
\begin{align}
H_{\mathrm{el}} &= \int d^3x ~\psi^{\dagger}(\mathbf{x})\left(-\frac{\hbar^2\nabla^2}{2m} + v_{\mathrm{ext}}(\mathbf{x}) \right)\psi(\mathbf{x}) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(\mathbf{x})\psi^{\dagger}(\mathbf{x}')V(\mathbf{x}-\mathbf{x}')\psi(\mathbf{x}')\psi(\mathbf{x}),
\end{align}
where $v_{\mathrm{ext}}$ is the one-body external potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
\begin{figure}[t]
\includegraphics[width=8.5cm]{conceptfigure.pdf}
\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons. (Right) Quasiparticle description of the coupled system as a factorizable state of an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
\label{fig:ansatz}
\end{figure}
Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$ renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{\mathrm{em}} = \frac{\epsilon_0}{2}\int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity.

@@

We capture the effect of correlations perturbatively. For the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{\infty}\sum_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
\label{eq:e-shift}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, and $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs. (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq. (\ref{eq:e-shift}). By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found. \textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq. (\ref{eq:e-shift}). These equations take into account the spatially varying wavefunctions to the spatially varying mode functions, just like Eqs. (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and therefore do not assume the dipole approximation.}
%\begin{figure}[t]
%\includegraphics[width=8.4cm]{figure3new.pdf}
%\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
%\label{fig:ansatz}
%\end{figure}

@@

Importantly, the reason perturbation theory fails for first excited state is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels. In contrast, no such near-resonance is found in the variational framework because the plasma term  in Eq. (\ref{eq:mhf-photon}) blue-shifts all of the photon frequencies. For the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV (from 0.62 eV),  far off-resonance from the bare emitter transition. The plasma term also strongly reduces the coupling between light and matter by a mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field (see \ref{fig:results}b). This is a  light-matter decoupling effect, which was proposed in Ref. ~\cite{liberato2014}, \textcolor{blue}{where, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, "effective field mode profiles" are obtained with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}

\textcolor{blue}{This light-matter decoupling is also reflected in Fig. ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle A\cdot p \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in SM). As shown in Fig. 3, numerical and variational methods capture a saturation and then decrease of this expectation value. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. The results of Fig. ~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}
